Practical Hilbert space approximate Bayesian Gaussian processes for probabilistic programming

### 2022-03-08 Paper/Manuscript_Stats&Comp/reply_letter_STCO.tex
edits to the responses to reviewer 2

The HSGP model is definitely useful in 1D and 2D, even for very wiggly functions, unless the number of data points is very small ($n \lesssim 300$) where the regular formulation of a GP is already reasonably fast itself.

It is difficult to give clear guidelines of when the application of the HSGP model can result to be impractical in other multidimensional cases, basically because it ultimately depends on the wigglyness of the function to be learned which is, in most of the cases, unknown. For that reason, in fact, the use of our diagnostics tool can be essential because it allows to iteratively fit the model using minimum computational requirements in each iteration. Observing how the requirements change in each iteration can be useful for the user to be able to recognize when the approximation is unfeasible to be applied. For $D>5$, even for very smooth functions, HSGPS definitely seem to be impractical. For $D<3$ it is widely useful and significantly faster than the exact GP even for very wiggly functions.

@@

Authors are really grateful to the reviewer for their valuable comments which have been very helpful to improve the paper, and for her/his time.

@@

From the experiments, we have built numerical functions that characterize the relationships between the key parameters of the model which are the basis to perform diagnostics on the fitting. We have developed a diagnostic procedure to fit, diagnose and update parameters, based on iterative steps until an accurate solution is reached. Furthermore, this diagnostic tool provides the optimal values for the key parameters of the model that allow for minimizing the computational cost, which can be essential for very large data sets, multi-dimensional cases, or other problems that require repeated fitting such as cross-validation and simulated based calibration.

The diagnostic rules, the numerical functions that allow updating the parameters and the recommended iterative steps to fit and diagnose the model are extensive to the community and essential to make the model usable for practitioners.

All of this have been properly explained along Section 4 in the paper.

@@

We have mention in the paper how dimensionality affects the expected number of basis functions needed and that the effect of dimensionality is only through this.

The HSGP model is definitely useful in 1D and 2D, even for very wiggly functions, unless the number of data points is very small ($n \lesssim 300$) where the regular formulation of a GP is already fast. It is difficult to give clear guidelines of when the application of the HSGP model can result to be impractical in other multidimensional cases, basically because it ultimately depends on the wigglyness of the function to be learned which is, in most of the cases, unknown. For that reason, in fact, the use of our diagnostics tool can be essential because it allows iteratively to fit the model using minimum computational requirements in each iteration. Observing how the requirements change in each iteration can be useful for the user to be able to recognize when the approximation is unfeasible to be applied. For $D>5$ even for very smooth functions seems to be impractical.

The Leukemia data set case study is a 4D data set. We have applied applied the diagnostic tool on it and we have been able to easily discover the optimal values for $c$ and $m$ that allows for a significantly faster performance than using the exact GP. Without the guide of the diagnostic tool the user probably would had made several trials until finding out an accurate and practical solution, consuming a lot of requirements in this demanding 4D case.

We have added a new case study to Section 5 which consists in wiggly 2D and 3D simulated data sets. We have applied the diagnostic procedure on it. Even though the diagnostic was built for unidimensional kernels, we demonstrate its applicability and usefulness in multidimensional cases by diagnosing individually the different dimensions in each iteration. It allows finding the  optimal solution with the minimum requirements that can be essential to use the HSGP model in multidimensional cases. We have included in this new case study a brief discussion and presentation of additional results concerning computational cost in 2D, 3D and 4D.

As commented across the paper and in particular in Section 4.8, the method is limited in practice to the number of dimensions and the wigglyness of the function across the different dimensions. We have observed that it can be useful for $D<5$ conditioned to the wigglyness of the function. However, for $D=2$ it is still useful and significantly faster than the exact GP even for very wiggly functions. In Section 4.8 we give brief details on the performance of the model in multi-D cases. In this sense, the use of our recommended diagnostic allows to find the optimal values for $c$ and $m$ which allows for minimum computational requirements which can be essential on these multidimensional cases.

@@

\textbf{There is no comparison with other techniques (except a spline method) ! Please compare with, e.g. "MCMC for Variationally Sparse Gaussian Processes" ; although this is SVI, it appears directly applicable for the problems considered here. Please also compare with non MCMC variational methods - while not directly applicable to generic MCMC methods, these are the sttandard and it would be nice to see how they compare in runtime.}

The main comparison is the Hilbert space basis function approximation against the exact GP as a model. In all comparisons the inference conditional on the model is assumed to be accurate. We've used efficient dynamic HMC, the convergence diagnostics didn't reveal problems, and thus the accuracy of the inference is good. Comparisons of MCMC to variational inference approximations are out of scope of this paper. Some comparisons of MCMC and variational inference for basis function GPs can be found in

@@

Comparisons of the Hilbert space basis function approximations against other GP approximations are provided by Solin and Särkkä (2020), which is cited prominently in our paper.

>>>>

The HSGP model is definitely useful in 1D and 2D, even for very wiggly functions, unless the number of data points is very small ($n \lesssim 300$) where the regular formulation of a GP is already reasonably fast itself.
It is difficult to give clear guidelines of when the application of the HSGP model can result to be impractical in other multidimensional cases, basically because it ultimately depends on the wigglyness of the function to be learned which is, in most of the cases, unknown. For that reason, in fact, the use of our diagnostics tool can be essential because it allows to iteratively fit the model using minimum computational requirements in each iteration. Observing how the requirements change in each iteration can be useful for the user to be able to recognize when the approximation is unfeasible to be applied. For $D>5$, even for very smooth functions, HSGPS definitely seem to be impractical. For $D<3$ it is widely useful and significantly faster than the exact GP even for very wiggly functions.

@@

@@

In the revised version of the paper, we have substantially extended the experiments. On that basis, we have derived equations that characterize the relationships between the key hyperparameters of the HSGP model and  the goodness of the corresponding exact GP approximation. We have developed a procedure to diagnose the appropriateness of the hyperparameters and to update them based on iterative steps until an accurate approximation has been achieved. Furthermore, this diagnostic tool provides optimal values for those hyperparameters with respect to minimizing the computational cost subject to sufficient accuracy of the approximation. Minimizing computation cost is particularily important for very large data sets, multi-dimensional cases, or other problems that require repeated fitting, such as cross-validation and simulated based calibration. All of this new content is discussed in detail in Section 4 of the paper.

%The diagnostic rules, the numerical functions that allow updating the parameters and the recommended iterative steps to fit and diagnose the model are extensive to the community and essential to make the model usable for practitioners.

@@

Indeed, the main limitation of the HSGP approach is its bad scaling with increasing number of input dimensions. In the revised version, we now make this more explicit and discuss the implications in detail, in particular in Section 4.8.

The HSGP model is definitely highly useful in 1D and 2D, even for very wiggly functions, unless the number of data points is very small ($n \lesssim 300$) where the regular formulation of a GP is already reasonably fast. It is difficult to give clear guidelines of when the application of the HSGP model becomes impractical in multidimensional cases, because it also depends on the wigglyness of the function to be learned which is, in most of the cases, unknown. For that reason, the use of our diagnostic tool is highly beneficial to minimize computational requirements in order to achieve a good approximation. Observing how the requirements change in each iteration can be useful for the user to be able to recognize when the approximation is unfeasible to be applied. For $D>5$, even for very smooth functions, HSGPs indeed seem to be impractical and we now state this explicitely.

In the Leukemia data set case study, we present a 4D HSGP model. We have applied the diagnostic tool and have been able to easily discover the optimal values for $c$ and $m$ that allows for a significantly faster performance than using the exact GP. Without the guidance of the diagnostic tool, a user would have to go through several more trials until finding a sufficiently accurate solution, which would result in substantially increased human and computer time requirements.

%As commented across the paper and in particular in Section 4.8, the method is limited in practice to the number of dimensions and the wigglyness of the function across the different dimensions. We have observed that it can be useful for $D<5$ conditioned to the wigglyness of the function. However, for $D=2$ it is still useful and significantly faster than the exact GP even for very wiggly functions. In Section 4.8 we give brief details on the performance of the model in multi-D cases. In this sense, the use of our recommended diagnostic allows to find the optimal values for $c$ and $m$ which allows for minimum computational requirements which can be essential on these multidimensional cases.

@@

\textbf{There is no comparison with other techniques (except a spline method) ! Please compare with, e.g. "MCMC for Variationally Sparse Gaussian Processes" ; although this is SVI, it appears directly applicable for the problems considered here. Please also compare with non MCMC variational methods - while not directly applicable to generic MCMC methods, these are the standard and it would be nice to see how they compare in runtime.}

The main comparison that we focus on is the Hilbert space basis function approximation (our HSGP method) against the exact GP model. In all comparisons, the inference conditional on the model is assumed to be accurate. In other words, we compare the exact to an approximate model, using asymptotically correct inference. For this purpose, we have used efficient dynamic HMC, the MCMC convergence diagnostics did not reveal problems, and thus the accuracy of the inference itself can be considered good.

While we agree that other comparisons could interesting as well, for example MCMC vs. variational inference, they would blow up the already extensive scope of the paper. Some comparisons of MCMC and variational inference for basis function GPs can be found in

@@

Further comparisons of the Hilbert space basis function approximations against other GP approximations are provided by Solin and Särkkä (2020), which is cited prominently in our paper.
